This is an example of article title

### 2025-10-30 paper/paper.tex
second example

@@

Physical systems can often be described by ordinary differential equations on Lie groups.
For example, the state of a charged metal rod moving on a plane can be specified by its position, orientation, linear velocity and angular velocity.
Such state naturally belongs to a Lie group: a direct product of \(G_1 = \mathrm{SO}(2) \times \mathbb{R}^2\) and its Lie algebra, \(\mathfrak{g}_1\), denoted \(G = G_1 \times \mathfrak{g}_1\).

{\color{red}TODO: Describe the ODE and have 2-3 code blocks}

@@

    \begin{tikzpicture}
        \begin{axis}[
            grid=major,
            width=.5\textwidth,
            legend style={at={(0.95,0.5)}, anchor=east},
            unit vector ratio=1 1,
            xmin = -1, xmax = 8,
            ymin = -4.25, ymax = 2.25,
            title style={yshift=-.5\baselineskip,},
            title={Trajectories of a charged rod between two point charges},
        ] %
            \addplot[only marks, mark=*, mark size=2pt, color=TolMutedSand] coordinates {(-0.5, 0.0)};
            \addlegendentry{$q_1$};
            \addplot[only marks, mark=*, mark size=2pt, color=TolMutedOlive] coordinates {(0.5, 0.0)};
            \addlegendentry{$q_2$};
            % Curve and quiver+-
            \addplot[solid, thick, TolMutedBlue] table[col sep=comma, x=x, y=y] {data/ex2_rod_1_dense.csv};
            \addlegendentry{$\gamma_1$}
            % Quiver plot from CSV file
            \addplot[
                quiver={u=\thisrow{up}, v=\thisrow{vp}, scale arrows=\arrowscale}, -{Bar[width=4pt]}, TolMutedBlue, forget plot,
            ] table[col sep=comma] {data/ex2_rod_1_sparse.csv};
            \addplot[
                quiver={u=\thisrow{um}, v=\thisrow{vm}, scale arrows=\arrowscale}, -{Circle[length=2pt]}, TolMutedBlue, forget plot,
            ] table[col sep=comma] {data/ex2_rod_1_sparse.csv};
            % Plot the base points from CSV file
            \addplot[
                only marks, mark=*, mark size=.5pt, TolMutedBlue, forget plot,
            ] table[col sep=comma] {data/ex2_rod_1_sparse.csv};
            %
            %
            % Curve and quiver+-
            \addplot[solid, thick, TolMutedGreen] table[col sep=comma, x=x, y=y] {data/ex2_rod_2_dense.csv};
            \addlegendentry{$\gamma_2$}
            % Quiver plot from CSV file
            \addplot[
                quiver={u=\thisrow{up}, v=\thisrow{vp}, scale arrows=\arrowscale}, -{Bar[width=4pt]}, TolMutedGreen, forget plot,
            ] table[col sep=comma] {data/ex2_rod_2_sparse.csv};
            \addplot[
                quiver={u=\thisrow{um}, v=\thisrow{vm}, scale arrows=\arrowscale}, -{Circle[length=2pt]}, TolMutedGreen, forget plot,
            ] table[col sep=comma] {data/ex2_rod_2_sparse.csv};
            % Plot the base points from CSV file
            \addplot[
                only marks, mark=*, mark size=.5pt, TolMutedGreen, forget plot,
            ] table[col sep=comma] {data/ex2_rod_2_sparse.csv};
            %
            %
            % Curve and quiver+-
            \addplot[solid, thick, TolMutedPurple] table[col sep=comma, x=x, y=y] {data/ex2_rod_3_dense.csv};
            \addlegendentry{$\gamma_3$}
            % Quiver plot from CSV file
            \addplot[
                quiver={u=\thisrow{up}, v=\thisrow{vp}, scale arrows=\arrowscale}, -{Bar[width=4pt]}, TolMutedPurple, forget plot,
            ] table[col sep=comma] {data/ex2_rod_3_sparse.csv};
            \addplot[
                quiver={u=\thisrow{um}, v=\thisrow{vm}, scale arrows=\arrowscale}, -{Circle[length=2pt]}, TolMutedPurple, forget plot,
            ] table[col sep=comma] {data/ex2_rod_3_sparse.csv};
            % Plot the base points from CSV file
            \addplot[
                only marks, mark=*, mark size=.5pt, TolMutedPurple, forget plot,
            ] table[col sep=comma] {data/ex2_rod_3_sparse.csv};
        \end{axis}
    \end{tikzpicture}
    \caption{Three solution curves \(\gamma_1\) (blue), \(\gamma_2\) (green), and \(\gamma_3\) (purple) of a Lie group Heun method on the product of Lie groups \(\mathrm{SO}(2) \times \mathbb R^2\)
    modelling the movement within an electric field with two point charges located at
    $q_1$ (sand) and $q_2$ (olive). }
    \label{fig:Heun}
\end{figure}

For more details on solving differential equations on Lie groups in Julia, see
\verb|ManifoldDiffEq.jl|.

{\color{red}TODO: Do a Forward Euler simulation on a nice physics problem}

>>>>

\usepackage{siunitx}

@@

Physical systems can often be described by ordinary differential equations on Lie groups.
For example, the state of a charged metal rod moving on a plane can be specified by its position, orientation, linear velocity and angular velocity.
Such state naturally belongs to a Lie group: a direct product of \(G_1 = \mathrm{SO}(2) \times \mathbb{R}^2\) and its Lie algebra, \(\mathfrak{g}_1\), denoted \(G = G_1 \times \mathfrak{g}_1\).
The ordinary differential equation governing the dynamics of the system can be derived from the electrostatic approximation, in particular the Coulomb's law.
Electric point charges \(q_i, i \in {1, \dots, N}\) at positions \(R_i\) generate a static electric field \(E(p)\) at point \(p\).
The force and the moment of force affecting the rod of length \(L\), mass \(M\), moment of inertia \(I_M\) linear charge density \(\lambda\), with the center at position \(p\), orientation \(\theta\) and where \(u\) is the direction along the rod can be expressed as
\begin{align}
    M\ddot{p} &= \lambda \int_{-L/2}^{L/2} E(p + su) ds,  \\
    I_M \ddot{\theta} &= \int_{-L/2}^{L/2} \lambda s u_{\perp} \cdot E(p + su) ds,
\end{align}
where \(u_{\perp}\) is the vector perpendicular to \(u\).
This can be understood as an ODE on \(G\) and solved using Lie groups methods, such as the RKMK variant of the Heun method~\cite{IserlesMunthe-KaasNoersettZanna:2000}.

Listing~\ref{lst:rod-dynamics} shows functions that calculate the right hand side of the ODE \lstinline{E_field} is a function that calculates electric field strength from the point charges at the given point.
The function \lstinline{force_and_torque} calculates the force and torque affecting the rod, while \lstinline{dynamics} converts them to the Lie algebra element can be used by the solver.
\begin{lstlisting}[
    language = Julia,
    numbers=left,
    label={lst:rod-dynamics},
    caption={Dynamics of a rod affected by point charges.}
]
function force_and_torque(sp::SystemParameters, Fθ, r)
    u, u⊥ = eachrow(Fθ)
    F, τ = zeros(2), 0.0
    for s in range(-sp.L/2, sp.L/2; length=Nq)
        E = E_field(sp, r .+ s .* u)
        F .+= sp.λ * E * (sp.L/(Nq-1))
        τ += sp.λ * s * dot(u⊥, E) * (sp.L/(Nq-1))
    end
    return F, τ
end

function dynamics(state, sp::SystemParameters, t)
    p, dp = state.x
    F, tau = force_and_torque(sp, p.x...)
    a = F ./ sp.M
    coords = [tau / sp.Icm, a...]
    ddx = hat(se2, coords, ArrayPartition)
    return ArrayPartition(dp, ddx)
end
\end{lstlisting}

The RKMK Heun method is a second order method that can solve ordinary differential equations defined on Lie groups.
A single step of an RKMK-type Heun solver is implemented in~\ref{lst:heun}.
The main advantage of using a Lie group solver for this problem is that we do not have to consider parametrization of the Lie group.
The presented approach can be easily generalized to three spatial dimensions, where parametrization of rotations would be even more challenging numerically.

\begin{lstlisting}[
    language = Julia,
    numbers=left,
    label={lst:heun},
    caption={Step calculation for the RKMK Heun method.}
]
function integrator_step!(A::GroupAction,
    sp::SystemParameters, y, f, t, dt)
    Ie = Identity(A.group)
    F1 = (dt / 2) * f(y, sp, t)
    tmp = zero_vector(A.manifold, y)
    diff_group_apply!(A, tmp, Ie, y, F1)
    y2 = exp(A.manifold, y, tmp)
    F2 = dt * f(y2, sp, t + dt/2)
    diff_group_apply!(A, tmp, Ie, y, F2)
    exp!(base_manifold(A), y, y, tmp)
end
\end{lstlisting}

The presented method was applied to a case with two static charges: \SI{1e-6}{C} at position (\SI{-1}{m}, \SI{0}{m}) and \SI{-1e-6}{C} at position (\SI{1}{m}, \SI{0}{m}).
A rod of length \SI{0.67}{m} with charge density \SI{1e-6}{C/m} and mass \SI{1}{kg} started from three different initial positions.
The trajectories obtained in a simulation with time step \SI{0.001}{s} are presented in Figure~\ref{fig:Heun}.
Simulation times between \SI{30}{s} and \SI{135}{s} were chosen to have the rod remain relatively close to point charges.
Position and orientation of the rod is indicated at 20 equally spaced times, along with a densely sampled trajectory of the center of the rod.

@@

    \begin{tikzpicture}
        \begin{axis}[
            grid=major,
            width=.5\textwidth,
            legend style={at={(0.95,0.25)}, anchor=east},
            unit vector ratio=1 1,
            xmin = -2, xmax = 5,
            ymin = -3.5, ymax = 3.5,
            title style={yshift=-.5\baselineskip,},
            title={Trajectories of a charged rod between two point charges},
        ] %
            \addplot[only marks, mark=*, mark size=2pt, color=TolMutedSand] coordinates {(-1, 0.0)};
            \addlegendentry{$q_1$};
            \addplot[only marks, mark=*, mark size=2pt, color=TolMutedOlive] coordinates {(1, 0.0)};
            \addlegendentry{$q_2$};
            % Curve and quiver+-
            \addplot[solid, thick, TolMutedBlue] table[col sep=comma, x=x, y=y] {data/ex2_rod_1_dense.csv};
            \addlegendentry{$\gamma_1$}
            % Quiver plot from CSV file
            \addplot[
                quiver={u=\thisrow{up}, v=\thisrow{vp}, scale arrows=\arrowscale}, -{Bar[width=4pt]}, TolMutedBlue, forget plot,
            ] table[col sep=comma] {data/ex2_rod_1_sparse.csv};
            \addplot[
                quiver={u=\thisrow{um}, v=\thisrow{vm}, scale arrows=\arrowscale}, -{Circle[length=2pt]}, TolMutedBlue, forget plot,
            ] table[col sep=comma] {data/ex2_rod_1_sparse.csv};
            % Plot the base points from CSV file
            \addplot[
                only marks, mark=*, mark size=.5pt, TolMutedBlue, forget plot,
            ] table[col sep=comma] {data/ex2_rod_1_sparse.csv};
            %
            %
            % Curve and quiver+-
            \addplot[solid, thick, TolMutedGreen] table[col sep=comma, x=x, y=y] {data/ex2_rod_2_dense.csv};
            \addlegendentry{$\gamma_2$}
            % Quiver plot from CSV file
            \addplot[
                quiver={u=\thisrow{up}, v=\thisrow{vp}, scale arrows=\arrowscale}, -{Bar[width=4pt]}, TolMutedGreen, forget plot,
            ] table[col sep=comma] {data/ex2_rod_2_sparse.csv};
            \addplot[
                quiver={u=\thisrow{um}, v=\thisrow{vm}, scale arrows=\arrowscale}, -{Circle[length=2pt]}, TolMutedGreen, forget plot,
            ] table[col sep=comma] {data/ex2_rod_2_sparse.csv};
            % Plot the base points from CSV file
            \addplot[
                only marks, mark=*, mark size=.5pt, TolMutedGreen, forget plot,
            ] table[col sep=comma] {data/ex2_rod_2_sparse.csv};
            %
            %
            % Curve and quiver+-
            \addplot[solid, thick, TolMutedPurple] table[col sep=comma, x=x, y=y] {data/ex2_rod_3_dense.csv};
            \addlegendentry{$\gamma_3$}
            % Quiver plot from CSV file
            \addplot[
                quiver={u=\thisrow{up}, v=\thisrow{vp}, scale arrows=\arrowscale}, -{Bar[width=4pt]}, TolMutedPurple, forget plot,
            ] table[col sep=comma] {data/ex2_rod_3_sparse.csv};
            \addplot[
                quiver={u=\thisrow{um}, v=\thisrow{vm}, scale arrows=\arrowscale}, -{Circle[length=2pt]}, TolMutedPurple, forget plot,
            ] table[col sep=comma] {data/ex2_rod_3_sparse.csv};
            % Plot the base points from CSV file
            \addplot[
                only marks, mark=*, mark size=.5pt, TolMutedPurple, forget plot,
            ] table[col sep=comma] {data/ex2_rod_3_sparse.csv};
        \end{axis}
    \end{tikzpicture}
    \caption{Three solution curves \(\gamma_1\) (blue), \(\gamma_2\) (green), and \(\gamma_3\) (purple) of a Lie group Heun method on the product of Lie groups \(\mathrm{SO}(2) \times \mathbb R^2\)
    modelling the movement within an electric field with two point charges located at
    $q_1$ (sand) and $q_2$ (olive). }
    \label{fig:Heun}
\end{figure}

For more advanced methods of solving differential equations on Lie groups in Julia, see
\verb|ManifoldDiffEq.jl|.
